Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-07-02 icfp24paper/paper.tex
fix
acknowledgements

\title{Closure-Free Functional Programming in a Two-Level Type Theory}
%% \titlenote{}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vb\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. This results from the dictionary-passing elaboration of
type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
produce additional closures.

@@

\section*{Acknowledgements}
I thank the anonymous reviewers and Sebastian Graf, for their comments and suggestions
on draft versions of this paper.

@@

In this appendix we describe the semantics of CFTT, from which the unstaging
algorithm is obtained, and also describe the soundness and stability of the
algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
definitions from there by default, and focus on differences and extensions.

@@

Next, we give a logical relation interpretation for $\vW$. We inductively define
a relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
the function type, which unfolds to pointwise preservation of relations. We also
define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w$ is a $\vsup$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. This definition
also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.

>>>>

\title{Closure-Free Functional Programming in a Two-Level Type Theory}

@@

Modern functional programming supports many convenient abstractions. These often
come with significant runtime overheads. Sometimes the overheads are acceptable,
but in other cases compiler optimization is crucial. Monads in Haskell is an
example for the latter. Even the $\Reader$ monad, which is one of the simplest
in terms of implementation, yields large overheads when compiled without
optimizations. Consider the following:
\begin{alignat*}{3}
  &\vf :: \Int \to \Reader\,\Bool\,\Int \\
  &\vf\,\vx = \mdo\{\vb \fro \msf{ask}; \bif \vx\, \bthen \return (\vx + 10)\, \belse \return (\vx + 20)\}
\end{alignat*}
With optimizations enabled, GHC compiles this roughly to the code below:
\begin{alignat*}{3}
  &\vf :: \Int \to \Bool \to \Int \\
  &\vf = \lam \vx\,\vb.\, \bif \vb\, \bthen \vx + 10\, \belse \vx + 20
\end{alignat*}
Without optimizations we roughly get:
\begin{alignat*}{3}
  &\vf = \lam \vx.\,(\bind)\,\msf{monadReader}\,(\msf{ask}\,\msf{monadReaderReader})\,(\lam \vb.\,\bif \vb\\
  &\ind \bthen\return\,\msf{monadReader}\,(\vx + 10)\\
  &\ind \belse\hspace{0.25em}\return\,\msf{monadReader}\,(\vx + 20))
\end{alignat*}
Here, $\msf{monadReader}$ and $\msf{monadReaderReader}$ are runtime
dictionaries, respectively for the $\msf{Monad}$ and $\msf{MonadReader}$
instances, and, for example, $(\bind)\,\msf{monadReader}$ is a field projection
from the dictionary. This results from the dictionary-passing elaboration of
type classes \cite{DBLP:conf/popl/WadlerB89}. We get a runtime closure from the
$\lam \vb.\,...$ function, and $(\bind)$, $\msf{ask}$ and $\mbf{return}$ also
produce additional closures.

@@

\section*{Acknowledgements}
I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
comments and suggestions on draft versions of this paper. This work was supported
by grant 2019.0116 of the Knut and Allice Wallenberg Foundation.

@@

In this appendix we describe the semantics of CFTT, from which the unstaging
algorithm is obtained, and also describe the soundness and completeness of the
algorithm. We borrow the setup from \cite{staged2ltt}; we use terminology and
definitions from there by default, and focus on differences and extensions.

@@

Next, we give a logical relation interpretation for $\vW$. We inductively define
a relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
the function type, which unfolds to pointwise preservation of relations. We also
define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w$ is a $\vsup$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma)$ and $\re\,(\welim\,P\,s\,w)\,\gamma')$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. This definition
also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.
